He-Man: A domain-specific language for lightweight concurrency

### 2012-03-06 paper/paper.tex
Added citations; partly cleaned up backend section.

\documentclass[preprint,10pt]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\usepackage{verbatim}

\usepackage{listings}
\lstset{
  language=Haskell,
  basicstyle=\small\ttfamily,
  frame=trbl,
  basewidth={0.5em,0.45em},
%  commentstyle=\ttfamily\color{ForestGreen},
  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
            {\\\\}{{\char`\\\char`\\}}1
            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
            {>>}{{>>}}2 {>>=}{{>>=}}2
            {|}{{$\mid$}}1
}

\lstnewenvironment{egcode}
  {\lstset{frame=TRBL}}
  {\lstset{frame=trbl}}

\lstnewenvironment{code}
  {\lstset{frame=}}
  {}

\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

@@

\begin{figure}[ht]
\centering
\begin{code}
child_code = declare_thread [("child_fd", Int)] $
  \[child_fd] -> do
  ev <- setup_connection child_fd
  buf <- new_buf bufsize
  while 1 $ do
    amt_read <- do_read ev buf bufsize
    ifthen (amt_read .== 0) exit
    amt_written <- full_write ev buf amt_read
    ifthen (amt_written .< amt_read) exit
\end{code}
\caption{He-Man example from echo server.}
\label{fig:code-ex}
\end{figure}

\begin{figure}[ht]
\centering
\begin{code}
do_nb_action :: IntE -> EventE -> Prog IntE
                -> Prog IntE
do_nb_action mode e action = do
  res <- var "result" Int (-1)
  while (res .< 0) $ do
    set_nb_event_mode e mode
    wait e
    res .=. action
  return res

accept (fd, e) =
  mk_fd <$> do_nb_action kEVENT_RD e (sock_accept fd)
do_read (fd, e) buf size =
  do_nb_action kEVENT_RD e (sock_read fd buf size)
do_write (fd, e) buf size =
  do_nb_action kEVENT_WR e (sock_write fd buf size)
\end{code}
\caption{Implementation of ``blocking'' network calls.}
\label{fig:generic}
\end{figure}

@@

\begin{figure}[ht]
\centering
\bnf{
\i{Thread}
\p{(\i{thread\_label}, [(\i{Var}, \i{Type})])}
\i{Block}
\p{(\i{label}, \i{thread\_label}, [\i{Stmt}], \i{Tail})}
\i{Tail}
\p{\b{Goto} \i{label}}
\a{\b{GotoWait} \i{label}}
\a{\b{If} \i{Expr} \i{Tail} \i{Tail}}
\a{\b{Exit}}
\i{Stmt}
\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
\a{\b{Assign} \i{Expr} \i{Expr}}
\a{\b{Exp} \i{Expr}}
\a{\b{Spawn} \i{label} [((\i{Var}, \i{Type}), \i{Expr})]}
}
\caption{Back-end abstract syntax.}
\label{fig:back-syntax}
\end{figure}

The next several passes convert our front-end grammar to a back-end grammar,
shown in Figure \ref{fig:back-syntax}, which closely resembles the features of
the C code that we subsequently generate. Most significantly, a program now
consists of a list of \i{Block}s of code, each of which belongs to a \i{Thread},
and a list of \i{Thread}s. \i{Expr}s remain unchanged from the front-end, while
all control flow is delegated to the \i{Tail} position of a \i{Block} of code.

As the grammar makes clear, the most important task for our back-end is to
convert our linear front-end code into a collection of states appropriate for an
event loop. This occurs in our \t{flattenPrgm} pass, which passes through the
program in reverse order with a \t{RWS} monad, converting it to uniquely-named
blocks whose only control-flow consists of tail-position jumps to other block
labels. Here, the \t{State} maintains the next block number, the \t{Writer}
collects lists of \i{Block}s and \i{Thread}s, and the \t{Reader} maintains the
current thread number (defined as the first block in the current thread).

As an illustrative example, consider encountering an \b{If} statement in our
backwards pass, where we have access to the translation of the later statements,
as well as the earlier statements. Morally, we would like to translate this to a
conditional jump to the translation of either the first or second arm, each of
which is terminated by a jump to the later statements. (Note that
\t{flattenPrgm} leaves all optimization to the next pass.)

We package up the later statements in a new block, and then translate the
preceding statements, terminated by a tail-position conditional \b{If}. We also
translate each arm of the original \b{If}, terminated by a tail-position
\b{Goto} to the block containing the later statements. Importantly, all three of
these recursive translations may generate new blocks, which will be collected
automatically by the \t{Writer} monad.

Because blocking I/O may only happen in the main loop, we also terminate a block
after any blocking I/O call, by inserting a \b{GotoWait} to the block containing
the later statements. \b{GotoWait} is identical to \b{Goto} except to signal to
later passes that the thread \emph{must} yield to the main loop at this
location, and thus the goto must not be optimized away.

This pass generates a correct state machine, albeit one which contains many
blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
pass, which removes these empty blocks, redirecting all \b{Goto}s to their
ultimate destinations, and allowing nested \b{If}s in tail position. Nested
\b{If}s occur frequently in generated code, for example, immediately before an
\t{if} statement at the end of a \t{while} loop, when the code might next jump
inside the \t{if} statement, to the top of the \t{while} loop, or after the
\t{while} loop, depending on multiple tests. Although this pass is, strictly
speaking, an optimization, it is critical for obtaining high-quality output.

At this point, the program's control flow is correct, but many blocks will
contain uses of variables which have been declared in previous blocks. The
\t{collectFrees} pass traverses each block, collecting the variables used but
not declared in each block. (Our front-end ensures that a variable is only ever
declared in one location, so there is no ambiguity.) The local variables of each
thread are the union of the used-not-declared variables in each of its blocks.
We save this list of local variables in the appropriate \i{Thread} container.

Note that we could simply mark all declared variables as thread-local, but
variables which are only used in same block as their declaration need not be
retained outside that block. We want to minimize the number of thread-local
variables, because this directly impacts the memory overhead of each thread.

>>>>

\documentclass[preprint,10pt]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\usepackage{verbatim}
\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

@@

\begin{figure}[ht]
\centering
\begin{smallverbatim}
child_code = declare_thread [("child_fd", Int)] $
  \[child_fd] -> do
  ev <- setup_connection child_fd
  buf <- new_buf bufsize
  while 1 $ do
    amt_read <- do_read ev buf bufsize
    ifthen (amt_read .== 0) exit
    amt_written <- full_write ev buf amt_read
    ifthen (amt_written .< amt_read) exit
\end{smallverbatim}
\caption{He-Man example from echo server.}
\label{fig:code-ex}
\end{figure}

\begin{figure}[ht]
\centering
\begin{smallverbatim}
do_nb_action :: IntE -> EventE -> Prog IntE
                -> Prog IntE
do_nb_action mode e action = do
  res <- var "result" Int (-1)
  while (res .< 0) $ do
    set_nb_event_mode e mode
    wait e
    res .=. action
  return res

accept (fd, e) =
  mk_fd <$> do_nb_action kEVENT_RD e (sock_accept fd)
do_read (fd, e) buf size =
  do_nb_action kEVENT_RD e (sock_read fd buf size)
do_write (fd, e) buf size =
  do_nb_action kEVENT_WR e (sock_write fd buf size)
\end{smallverbatim}
\caption{Implementation of ``blocking'' network calls.}
\label{fig:generic}
\end{figure}

@@

\begin{figure}[ht]
\centering
\bnf{
\i{Thread}
\p{(\i{thread\_label}, [(\i{Var}, \i{Type})])}
\i{Block}
\p{(\i{label}, \i{thread\_label}, [\i{Stmt}], \i{Tail})}
\i{Tail}
\p{\b{Goto} \i{label}}
\a{\b{GotoWait} \i{Expr} \i{label}}
\a{\b{If} \i{Expr} ([\i{Stmt}], \i{Tail}) ([\i{Stmt}], \i{Tail})}
\a{\b{Exit}}
\i{Stmt}
\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
\a{\b{Assign} \i{Expr} \i{Expr}}
\a{\b{Exp} \i{Expr}}
\a{\b{Spawn} \i{label} [((\i{Var}, \i{Type}), \i{Expr})]}
}
\caption{Back-end abstract syntax.}
\label{fig:back-syntax}
\end{figure}

The next several passes convert our front-end grammar to a back-end grammar,
shown in Figure \ref{fig:back-syntax}, which closely resembles the C code we
ultimately generate. In particular, a back-end program is a list of code
\i{Block}s, each belonging to one of a list of \i{Thread}s. \i{Expr}s remain
unchanged from the front-end, but critically, control flow is relegated to
\i{Tail} position in each block.

Our back-end's main task is to convert our linear front-end code into a state
machine of code blocks. This occurs in our \t{flattenPrgm} pass, which passes
through program \i{Stmt}s in reverse order, generating uniquely-named blocks
whose only control-flow consists of (non-nested) tail-position jumps to other
blocks. Threads are identified with their first block.
% TODO discuss the transformations

%As an illustrative example, consider encountering an \b{If} statement in our
%backwards pass, where we have access to the translation of the later
%statements, as well as the earlier statements. Morally, we would like to
%translate this to a conditional jump to the translation of either the first or
%second arm, each of which is terminated by a jump to the later statements.
%(Note that \t{flattenPrgm} leaves all optimization to the next pass.)

%We package up the later statements in a new block, and then translate the
%preceding statements, terminated by a tail-position conditional \b{If}. We also
%translate each arm of the original \b{If}, terminated by a tail-position
%\b{Goto} to the block containing the later statements. Importantly, all three
%of these recursive translations may generate new blocks, which will be
%collected automatically by the \t{Writer} monad.

%Because blocking I/O may only happen in the main loop, we also terminate a
%block after any blocking I/O call, by inserting a \b{GotoWait} to the block
%containing the later statements. \b{GotoWait} is identical to \b{Goto} except
%to signal to later passes that the thread \emph{must} yield to the main loop at
%this location, and thus the goto must not be optimized away.
% TODO for concurrency-related reasons

This pass generates a correct, albeit unnecessarily large, state machine with
many empty blocks. Over the next few passes, our goal is to shrink this state
machine as much as possible, because yielding to the main loop is the only
overhead in the He-Man system. Although these passes are, strictly speaking,
optimizations, they are very important for performance.

Our \t{simplifyJumps} pass traverses the program collecting empty blocks,
redirecting control flow to its ultimate destination. In particular, we often
introduce nested \b{If}s in tail-position, for example, before an \t{if}
statement at the end of a \t{while} loop, where control resumes in one of three
blocks.

At this point, we need to eliminate spurious control flow edges from our state
machine in order to enable further block elimination. To do this, we perform
block-local constant propagation and folding, and dead code elimination. At face
value, it is not important for us to optimize straight-line code: our output is
further optimized by a C compiler, and execution time is dominated by I/O, not
computation. However, simple constant folding eliminates many \t{if} statements
%TODO examples, justification

Once our 
% TODO fuseBlocks pass

At this point, many blocks refer to variables declared in other blocks. The
\t{collectFrees} pass traverses each block, collecting the variables each thread
must close over. Because our front-end ensures each variable declaration is
unique, we need not consider control flow; the set of thread-local variables is
merely the union of the used-not-declared variables in each of its blocks.  We
save this list of local variables in the appropriate \i{Thread} container. (Of
course, it would be adequate to mark all variables thread-local, but this would
negatively impact our per-thread memory overhead.)
